---
title: "Luna Dragen VSP"
author: "Pender, J."
date: "Last updated on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  fig_caption: yes
  pdf_document: null
editor_options:
  chunk_output_type: console
---


# Setting up working directory, packages
```{r}
setwd("~/sequencing-analysis/sequoia/outputs-sequoia/microorganisms_and_amr")
setwd("~/github/sequoia-dragen/outputs-sequoia/microorganisms_and_amr")
rm(list = ls())

library("jsonlite")
library("dplyr")
library("ggplot2")
library("purrr")
library("patchwork")
library("tidyverse")
library("stringr")

```


# Flatten to named row function for coverting sublist microorganisms or amr markers from a file into a data frame
```{r}  
# recursive function that takes a nested list and fills out the fields for the list of hits in a sample
flatten_to_named_row <- function(x, field.prefix = NULL) {
  result <- list() # initialize place to collect 

  # loop over each field in the list
  for (name in names(x)) {
    full_name <- if (!is.null(field.prefix)) paste(field.prefix, name, sep = ".") else name
    val <- x[[name]]

    if (is.null(val)) {
      result[[full_name]] <- NA

      # handles cases such as '0.988, "Group 2"'
    } else if (is.atomic(val) && length(val) == 1) {
      # Single value → keep
      result[[full_name]] <- val

      # joins things like character[8] c("A", "B", "C")) into one string
    } else if (is.atomic(val)) {
      # Vector (no nesting) → collapse into single string
      result[[full_name]] <- paste(val, collapse = ", ")

      # handles sublists with atomic values only
    } else if (is.list(val) && all(sapply(val, function(v) is.atomic(v) || is.null(v)))) {
      # if field is a list of atomic values with no sublists, loop through each subfield and create compound name like associatedAmrMarkers.detected
      for (subname in names(val)) {
        subval <- val[[subname]]
        sub_full_name <- paste(full_name, subname, sep = ".")
        if (is.null(subval)) {
          result[[sub_full_name]] <- NA
          # if the subvalue is scalar, store as is:
        } else if (length(subval) == 1) {
          result[[sub_full_name]] <- subval
          # if it's a vector, collapse it into a string:
        } else {
          result[[sub_full_name]] <- paste(subval, collapse = ", ")
        }
      }

      # If the value is in a deeper nested list, handle it recursively.
    } else {
      # Deeper nested list → recurse
      nested <- flatten_to_named_row(val, field.prefix = full_name)
      result <- c(result, nested)
    }
  }
  # Convert the list into a 1-row data frame
  as.data.frame(result, stringsAsFactors = FALSE)
}

```

# Load json files
```{r}

json_files_urinary <- list.files("../../inputs-sequoia/sequoia-upip/", pattern = "*.json", full.names = TRUE)
df_urinary <- data.frame()

json_files_respiratory <- list.files("../../inputs-sequoia/sequoia-rpip/", pattern = "*.json", full.names = TRUE)
df_respiratory <- data.frame()

# json_files_urinary.amrs.only <- list.files("../../inputs-sequoia/sequoia-upip-amr-markers-only/", pattern = "*.json", full.names = TRUE)
# df_urinary <- data.frame()

```

# single file troubleshooting
```{r}
file <- "../../inputs-sequoia/sequoia-upip/0505-MOD-SOL.UPIP.report.json"
json_data <- fromJSON(file, simplifyDataFrame = FALSE)

microorganisms <- json_data.upip$targetReport$microorganisms
amrMarkers <- json_data.rpip$targetReport$amrMarkers

# apply the function to each list in the file
df.microorganisms.single <- bind_rows(lapply(microorganisms, flatten_to_named_row))
df.amr.markers.single <- bind_rows(lapply(amrMarkers, flatten_to_named_row))

df.microorganisms.single$sample <- sub("\\..*", "", basename(file))
df.amr.markers.single$sample <- sub("\\..*", "", basename(file))

df.microorganisms.single$type <- case_when(
      grepl("(^|-)INF$", df.microorganisms.single$sample) ~ "Influent",
      grepl("(^|-)SOL$", df.microorganisms.single$sample) ~ "Solids",
      TRUE ~ "Other"
)  
df.amr.markers.single$type <- case_when(
      grepl("(^|-)INF$", df.amr.markers.single$sample) ~ "Influent",
      grepl("(^|-)SOL$", df.amr.markers.single$sample) ~ "Solids",
      TRUE ~ "Other" 
)  

df.microorganisms.single$relabund.rpkm <- as.numeric(df.microorganisms.single$rpkm) / sum(as.numeric(df.microorganisms.single$rpkm), na.rm = TRUE)  
df.amr.markers.single$relabund.rpkm <- as.numeric(df.amr.markers.single$rpkm) / sum(as.numeric(df.amr.markers.single$rpkm), na.rm = TRUE)

```

# Create dataframes
```{r}

process_json <- function(json_files, target = c("amrMarkers", "microorganisms")) {
  target <- match.arg(target)  # ensures valid input
  df_all <- data.frame()

  for (file in json_files) {
    json_data <- fromJSON(file, simplifyDataFrame = FALSE)
    sample_name <- sub("\\..*", "", basename(file))

    # Extract the right target list from the JSON
    records <- switch(
      target,
      amrMarkers = json_data$targetReport$amrMarkers,
      microorganisms = json_data$targetReport$microorganisms
    )

    # Skip if empty or NULL
    if (is.null(records) || length(records) == 0) {
      # Include a placeholder row if the record is empty
      df <- data.frame(sample = sample_name)
    } else {
      df <- bind_rows(lapply(records, flatten_to_named_row))
      df$sample <- sample_name
    }

    df_all <- bind_rows(df_all, df)
  }

  if ("sample" %in% names(df_all)) {
    # Move 'sample' to the first column if it exists
    df_all <- df_all[, c("sample", setdiff(names(df_all), "sample"))]
  }
  
  return(df_all)
}

df_amr.markers.urinary <- process_json(json_files_urinary, target = "amrMarkers")
df_micro.urinary <- process_json(json_files_urinary, target = "microorganisms")

df_amr.markers.respiratory <- process_json(json_files_respiratory, target = "amrMarkers")
df_micro.respiratory <- process_json(json_files_respiratory, target = "microorganisms")

```

# Partition samples function
```{r}
partition_samples <- function(df) {
  
  controls <- c("BLANK-1", "BLANK-2", "VIRCELL-ADV41", "VIRCELL-H1", "ZEPTO-ADV1", "ZEPTO-H1")
  nonsamples <- c("AACTTATCCT-ACGCCAGTAC", "CCAATGATAC-CGGCACACTC", "CCATGGTATA-CTCCGCGAGA",
                        "CGATCTGTGA-AGTCTTCCTA", "GTAACAATCT-GTATACAGAG", "GTAGGCGAGC-TGCGCTCCTA",
                        "GTGGACAAGT-AAGGCCACGG", "TCTCAATACC-CGCGAACGTC", "TCTTGTCGGC-ACACAGGTGG",
                        "TGCAAGATAA-CCGTCCTCAA", "TTAGACCATG-CTTCCTAGGA",
                        "Undetermined-lane2", "Undetermined-lane3"
                        )
  
  df_controls <- df[df$sample %in% controls, ]
  df_nonsamples <- df[df$sample %in% nonsamples, ]
  df_sample.spike  <- df %>% filter(grepl("^0718-WOD-INF", sample))
  
  # excludes controls and six-replicate 3-spike sample
  df_samples <- df %>%
    filter(
      !(sample %in% controls) &       # Exclude controls
      !(sample %in% nonsamples) &
      !grepl("^0718-WOD-INF", sample) # Exclude anything starting with "0718-WOD-INF"
    )
  
  # Grabs samples names that end with INF or SOL. Does not include 0718-WOD-INF- replicates
  df_influent <- df %>% filter(grepl("-INF$", sample))
  df_solids <- df %>% filter(grepl("-SOL$", sample))
  
  return(list(
      controls = df_controls,
      nonsamples = df_nonsamples,
      sample.spike = df_sample.spike,
      samples = df_samples,
      influent = df_influent,
      solids = df_solids
  ))
}

partitioned_amr.markers.urinary <- partition_samples(df_amr.markers.urinary)
partitioned_micro.urinary <- partition_samples(df_micro.urinary)

partitioned_amr.markers.respiratory <- partition_samples(df_amr.markers.respiratory)
partitioned_micro.respiratory <- partition_samples(df_micro.respiratory)


```

# Separate partitions
```{r}
df_amr.markers.urinary_controls <- partitioned_amr.markers.urinary$controls
df_amr.markers.urinary_nonsamples <- partitioned_amr.markers.urinary$nonsamples
df_amr.markers.urinary_sample.spike <- partitioned_amr.markers.urinary$sample.spike
df_amr.markers.urinary_samples <- partitioned_amr.markers.urinary$samples
df_amr.markers.urinary_influent <- partitioned_amr.markers.urinary$influent
df_amr.markers.urinary_solids <- partitioned_amr.markers.urinary$solids

df_micro.urinary_controls <- partitioned_micro.urinary$controls
df_micro.urinary_nonsamples <- partitioned_micro.urinary$nonsamples
df_micro.urinary_sample.spike <- partitioned_micro.urinary$sample.spike
df_micro.urinary_samples <- partitioned_micro.urinary$samples
df_micro.urinary_influent <- partitioned_micro.urinary$influent
df_micro.urinary_solids <- partitioned_micro.urinary$solids

df_amr.markers.respiratory_controls <- partitioned_amr.markers.urinary$controls
df_amr.markers.respiratory_nonsamples <- partitioned_amr.markers.urinary$nonsamples
df_amr.markers.respiratory_sample.spike <- partitioned_amr.markers.urinary$sample.spike
df_amr.markers.respiratory_samples <- partitioned_amr.markers.urinary$samples
df_amr.markers.respiratory_influent <- partitioned_amr.markers.urinary$influent
df_amr.markers.respiratory_solids <- partitioned_amr.markers.urinary$solids

df_micro.respiratory_controls <- partitioned_micro.respiratory$controls
df_micro.respiratory_nonsamples <- partitioned_micro.respiratory$nonsamples
df_micro.respiratory_sample.spike <- partitioned_micro.respiratory$sample.spike
df_micro.respiratory_samples <- partitioned_micro.respiratory$samples
df_micro.respiratory_influent <- partitioned_micro.respiratory$influent
df_micro.respiratory_solids <- partitioned_micro.respiratory$solids

```

# process data frame - new
```{r}
process_data_frame <- function(df, remove_spike_OC43 = FALSE) {

  all.combinations <- expand_grid(
    sample = unique(df$sample),
    name = unique(df$name)
  )
  
  all.combinations.filled <- all.combinations %>%
    left_join(df, by = c("sample", "name"))
  
  # Get rid of placeholder NA row from earlier for samples that didn't have hits
  all.combinations.filled <- all.combinations.filled %>% filter(!is.na(name))
  
  df_zeros <- all.combinations.filled %>%
    mutate(
      alignedReadCount = replace_na(alignedReadCount, 0),
      rpkm = replace_na(rpkm, 0)
    )
  
  df_zeros$type <- case_when(
        grepl("(^|-)INF$", df_zeros$sample) ~ "Influent",
        grepl("(^|-)SOL$", df_zeros$sample) ~ "Solids",
        TRUE ~ "Other"
      )
  
  # Take out type from the sample name (0505-MOD-SOL -> 0505-MOD)
  df_zeros$sample <- gsub("-(SOL|INF)", "", df_zeros$sample)
  
  df_complete <- df_zeros %>%
    filter(type %in% c("Solids", "Influent", "Other")) %>%
    mutate(type = factor(type, levels = c("Solids", "Influent", "Other")))
  
   if (remove_spike_OC43) {
    df_complete <- df_complete %>%
      filter(!name %in% c("Human coronavirus OC43 (HCoV_OC43)"))
  }
  
  df_complete <- df_complete %>%
  group_by(sample) %>%
  mutate(
    relabund.rpkm = if (sum(rpkm) == 0) 0 else rpkm / sum(rpkm),
    percent.rpkm = round(relabund.rpkm * 100, 2)
  ) %>%
  ungroup()
  
  df_complete$name <- factor(df_complete$name, levels = df_complete %>%
                           group_by(name) %>%
                           summarise(total.relabund.rpkm = sum(relabund.rpkm)) %>%
                           arrange(total.relabund.rpkm) %>%
                           pull(name))
  
  # adds location from sample name, capturing everything after the first '-' and up to the next '-' or end of string
  var_name <- deparse(substitute(df))
  df_complete$location <- if (grepl("(_samples|_sample\\.spike|_solids|_influent)$", var_name)) {
    sub("^[^\\-]+-([^\\-]+).*", "\\1", df_complete$sample)
  } else {
    NA
  }
  
  return(df_complete)
}

```

# process data frame for each partition
```{r}
df_amr.markers.urinary_controls_processed <- process_data_frame(df_amr.markers.urinary_controls)
df_amr.markers.urinary_nonsamples_processed <- process_data_frame(df_amr.markers.urinary_nonsamples)
df_amr.markers.urinary_sample.spike_processed <- process_data_frame(df_amr.markers.urinary_sample.spike)
df_amr.markers.urinary_samples_processed <- process_data_frame(df_amr.markers.urinary_samples, remove_spike_OC43 = TRUE)
df_amr.markers.urinary_influent_processed <- process_data_frame(df_amr.markers.urinary_influent, remove_spike_OC43 = TRUE)
df_amr.markers.urinary_solids_processed <- process_data_frame(df_amr.markers.urinary_solids, remove_spike_OC43 = TRUE)

df_micro.urinary_controls_processed <- process_data_frame(df_micro.urinary_controls)
df_micro.urinary_nonsamples_processed <- process_data_frame(df_micro.urinary_nonsamples)
df_micro.urinary_sample.spike_processed <- process_data_frame(df_micro.urinary_sample.spike)
df_micro.urinary_samples_processed <- process_data_frame(df_micro.urinary_samples, remove_spike_OC43 = TRUE)
df_micro.urinary_influent_processed <- process_data_frame(df_micro.urinary_influent, remove_spike_OC43 = TRUE)
df_micro.urinary_solids_processed <- process_data_frame(df_micro.urinary_solids, remove_spike_OC43 = TRUE)

df_amr.markers.respiratory_controls_processed <- process_data_frame(df_amr.markers.respiratory_controls)
df_amr.markers.respiratory_nonsamples_processed <- process_data_frame(df_amr.markers.respiratory_nonsamples)
df_amr.markers.respiratory_sample.spike_processed <- process_data_frame(df_amr.markers.respiratory_sample.spike)
df_amr.markers.respiratory_samples_processed <- process_data_frame(df_amr.markers.respiratory_samples, remove_spike_OC43 = TRUE)
df_amr.markers.respiratory_influent_processed <- process_data_frame(df_amr.markers.respiratory_influent, remove_spike_OC43 = TRUE)
df_amr.markers.respiratory_solids_processed <- process_data_frame(df_amr.markers.respiratory_solids, remove_spike_OC43 = TRUE)

df_micro.respiratory_controls_processed <- process_data_frame(df_micro.respiratory_controls)
df_micro.respiratory_nonsamples_processed <- process_data_frame(df_micro.respiratory_nonsamples)
df_micro.respiratory_sample.spike_processed <- process_data_frame(df_micro.respiratory_sample.spike)
df_micro.respiratory_samples_processed <- process_data_frame(df_micro.respiratory_samples, remove_spike_OC43 = TRUE)
df_micro.respiratory_influent_processed <- process_data_frame(df_micro.respiratory_influent, remove_spike_OC43 = TRUE)
df_micro.respiratory_solids_processed <- process_data_frame(df_micro.respiratory_solids, remove_spike_OC43 = TRUE)
```

# Function to consolidate samples by type or location
```{r}
consolidate <- function(df, consolidation.level = c("type", "location")) {
  consolidation.level <- match.arg(consolidation.level)
  prefix <- sub("_processed$", "", sub("^df_", "", deparse(substitute(df))))

  # Group by the chosen level and name
  df_avg.relabund <- df %>%
    group_by(.data[[consolidation.level]], name) %>%
    summarize(relabund.rpkm.mean = mean(relabund.rpkm, na.rm = TRUE), .groups = "drop") %>%
    group_by(.data[[consolidation.level]]) %>%
    mutate(relabund.rpkm.mean.normalized = relabund.rpkm.mean / sum(relabund.rpkm.mean, na.rm = TRUE)) %>%
    ungroup()

  df_avg.rpkm <- df %>%
    group_by(.data[[consolidation.level]], name) %>%
    summarize(rpkm.mean = mean(rpkm, na.rm = TRUE), .groups = "drop")

  # Join the two summaries
  df_combined <- left_join(
    df_avg.relabund,
    df_avg.rpkm,
    by = c(consolidation.level, "name")
  )

  return(df_combined)
}

# Summarize by location
df_amr.markers.urinary_influent_by_location <- consolidate(df_amr.markers.urinary_influent_processed, "location")
df_amr.markers.urinary_solids_by_location <- consolidate(df_amr.markers.urinary_solids_processed, "location")

df_micro.urinary_influent_by_location <- consolidate(df_micro.urinary_influent_processed, "location")
df_micro.urinary_solids_by_location <- consolidate(df_micro.urinary_solids_processed, "location")

df_amr.markers.respiratory_influent_by_location <- consolidate(df_amr.markers.respiratory_influent_processed, "location")
df_amr.markers.respiratory_solids_by_location <- consolidate(df_amr.markers.respiratory_solids_processed, "location")

df_micro.respiratory_influent_by_location <- consolidate(df_micro.respiratory_influent_processed, "location")
df_micro.respiratory_solids_by_location <- consolidate(df_micro.respiratory_solids_processed, "location")

# Summarize by type
df_amr.markers.urinary_samples_by_type <- consolidate(df_amr.markers.urinary_samples_processed, "type")
df_micro.urinary_samples_by_type <- consolidate(df_micro.urinary_samples_processed, "type")
df_amr.markers.respiratory_samples_by_type <- consolidate(df_amr.markers.respiratory_samples_processed, "type")
df_micro.respiratory_samples_by_type <- consolidate(df_micro.respiratory_samples_processed, "type")

```


# analysis
```{r}

df_urinary_unique <- df_urinary %>%
  distinct(Name, .keep_all = TRUE) %>%  # keep only first row of each Name
  select(Name, Drug, Microorganism)     # select columns of interest

write.csv(df_urinary_unique, "urinary_unique.csv", row.names = FALSE)

df_respiratory_unique <- df_respiratory %>%
  distinct(Name, .keep_all = TRUE) %>%  # keep only first row of each Name
  select(Name, Drug, Microorganism)     # select columns of interest

write.csv(df_respiratory_unique, "respiratory_unique.csv", row.names = FALSE)

```


# Non-consolidated abundance - controls, weird samples
```{r}
plot_abundance <- function(df, width, height, ncol)  {
  # Extract the prefix from the input df name
  prefix <- sub("_processed$", "", sub("^df_", "", deparse(substitute(df))))
  
  # Determine analysis and feature types from the prefix
  upip_rpip <- case_when(
    grepl("urinary", prefix)    ~ "Urinary",
    grepl("respiratory", prefix) ~ "Respiratory",
    TRUE                         ~ "<unknown>")
  micro_amr <- case_when(
    grepl("micro", prefix)        ~ "Microorganisms",
    grepl("amr\\.markers", prefix) ~ "AMR Markers",
    TRUE                           ~ "<unknown>")
  sample_type <- case_when(
    grepl("controls", prefix)       ~ "Controls",
    grepl("nonsamples", prefix)     ~ "Non-Samples",
    grepl("sample\\.spike", prefix) ~ "Tri-Spike Replicates",
    TRUE                            ~ "")
  fill_label    <- paste(upip_rpip, micro_amr)
  
  title_base <- paste(upip_rpip, micro_amr, "in", sample_type)
  ref_df <- paste0("(", prefix, ")")
  
  # Individual titles
  relabund_title <- paste("Relative Abundance of", title_base, "\n", ref_df)
  abundance_title <- paste("Raw Abundance of", title_base, "\n", ref_df)
  relabund_label <- "Relative Abundance"
  abundance_label <- "Raw Abundance"
  plot_title <- paste("Relative and Raw Abundance of", title_base, "\n", ref_df)

  # Base plot styling
  base_theme <- theme_minimal() +
    theme(
      axis.line.x = element_line(color = "black"),
      axis.line.y = element_line(color = "black"),
      panel.grid = element_blank(),
      axis.ticks.x = element_line(),
      axis.ticks.y = element_line(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      plot.margin = margin(10, 10, 10, 10)
    )
  
# Plot 1: Relative Abundance (relabund.rpkm)
  p_relabund <- ggplot(df, aes(x = sample, y = relabund.rpkm, fill = name)) +
    geom_bar(stat = "identity", color = "Dark Blue") +
    scale_y_continuous(limits = c(0, 1.01), expand = c(0, 0)) +
    scale_x_discrete(expand = c(0.09, 0.09)) +
    labs(
      title = relabund_title,
      x = sample_type,
      y = "Relative Abundance of RPKM",
      fill = fill_label
    ) +
    guides(fill = guide_legend(ncol = ncol)) +
    base_theme

  # Plot 2: Raw RPKM abundance
  p_rpkm <- ggplot(df, aes(x = sample, y = rpkm, fill = name)) +
    geom_bar(stat = "identity", color = "Dark Blue") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    scale_x_discrete(expand = c(0.09, 0.09)) +
    labs(
      title = abundance_title,
      x = sample_type,
      y = "RPKM",
      fill = fill_label
    ) +
    guides(fill = guide_legend(ncol = ncol)) +
    base_theme
  
  # Combining relabund and abundance plots
  # Remove legend from left plot
  p_relabund_no_legend <- p_relabund + theme(legend.position = "none") + labs(title = relabund_label)
  p_rpkm_with_legend <- p_rpkm + labs(title = abundance_label)

  p_combined <- (p_relabund_no_legend | p_rpkm_with_legend) +
  plot_layout(guides = "collect") +
    plot_annotation(
    title = plot_title,
    theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 10)))
  ) &
  theme(legend.position = "right")
  
  # Save plots in separate folders
  ggsave(file = file.path("nonconsolidated_graphs/1.relabund", paste0("plot_relabund_", prefix, ".pdf")), 
         plot = p_relabund, width = width, height = height)
  
  ggsave(file = file.path("nonconsolidated_graphs/2.rpkm", paste0("plot_rpkm_", prefix, ".pdf")), 
         plot = p_rpkm, width = width, height = height)
  
  ggsave(file = file.path("nonconsolidated_graphs/3.combined", paste0("plot_combined_", prefix, ".pdf")), 
         plot = p_combined, width = width * 1.5, height = height)

  # Return both plots as a named list
  return(list(
    relabund_plot = p_relabund,
    rpkm_plot = p_rpkm,
    combined_plot = p_combined
  ))
}

# Create directories if they don't already exist
dir.create("nonconsolidated_graphs/1.relabund", showWarnings = FALSE, recursive = TRUE)
dir.create("nonconsolidated_graphs/2.rpkm", showWarnings = FALSE, recursive = TRUE)
dir.create("nonconsolidated_graphs/3.combined", showWarnings = FALSE, recursive = TRUE)

plots_amr.markers.respiratory_controls <- plot_abundance(df_amr.markers.respiratory_controls_processed, w=11, h=6.5, n=1)
plots_amr.markers.respiratory_nonsamples <- plot_abundance(df_amr.markers.respiratory_nonsamples_processed, w=14, h=6.5, n=5)
plots_amr.markers.respiratory_sample.spike <- plot_abundance(df_amr.markers.respiratory_sample.spike_processed, w=11, h=6.5, n=5)

plots_amr.markers.urinary_controls <- plot_abundance(df_amr.markers.urinary_controls_processed, w=11, h=6.5, n=1)
plots_amr.markers.urinary_nonsamples <- plot_abundance(df_amr.markers.urinary_nonsamples_processed, w=14, h=6.5, n=5)
plots_amr.markers.urinary_sample.spike <- plot_abundance(df_amr.markers.urinary_sample.spike_processed, w=11, h=6.5, n=5)

plots_micro.respiratory_controls <- plot_abundance(df_micro.respiratory_controls_processed, w=11, h=6.5, n=1)
plots_micro.respiratory_nonsamples <- plot_abundance(df_micro.respiratory_nonsamples_processed, w=14, h=6.5, n=2)
plots_micro.respiratory_sample.spike <- plot_abundance(df_micro.respiratory_sample.spike_processed, w=11, h=6.5, n=1)

plots_micro.urinary_controls <- plot_abundance(df_micro.urinary_controls_processed, w=11, h=6.5, n=1)
plots_micro.urinary_nonsamples <- plot_abundance(df_micro.urinary_nonsamples_processed, w=14, h=6.5, n=2)
plots_micro.urinary_sample.spike <- plot_abundance(df_micro.urinary_sample.spike_processed, w=11, h=6.5, n=1)


# # View them
# plots_micro.respiratory_controls$relabund_plot
# plots_micro.respiratory_controls$rpkm_plot
# plots_micro.respiratory_controls$combined_plot

```


# Function to plot average consolidated abundance
```{r}
plot_consolidated_abundance <- function(df, consolidation.level = c("type", "location"), width, height, ncol)  {
  consolidation.level <- match.arg(consolidation.level)
  # Extract the prefix from the input df name
  prefix <- sub("_processed$", "", sub("^df_", "", deparse(substitute(df))))
  
  # Determine analysis and feature types from the prefix
  upip_rpip <- case_when(
    grepl("urinary", prefix)    ~ "Urinary",
    grepl("respiratory", prefix) ~ "Respiratory",
    TRUE                         ~ "<unknown>")
  micro_amr <- case_when(
    grepl("micro", prefix)        ~ "Microorganisms",
    grepl("amr\\.markers", prefix) ~ "AMR Markers",
    TRUE                           ~ "<unknown>")
  sample_type <- case_when(
    grepl("solids", prefix)   ~ "in Solids",
    grepl("influent", prefix) ~ "in Influent",
    TRUE                      ~ "")
  fill_label    <- paste(upip_rpip, micro_amr)
  
  title_base <- paste(upip_rpip, micro_amr, sample_type, "Consolidated to", tools::toTitleCase(consolidation.level))
  ref_df <- paste0("(", prefix, ")")
  
  # Individual titles
  relabund_title <- paste("Relative Abundance of", title_base, "\n", ref_df)
  abundance_title <- paste("Raw Abundance of", title_base, "\n", ref_df)
  relabund_label <- "Relative Abundance"
  abundance_label <- "Raw Abundance"
  plot_title <- paste("Relative and Raw Abundance of", title_base, "\n", ref_df)

  # Base plot styling
  base_theme <- theme_minimal() +
    theme(
      axis.line.x = element_line(color = "black"),
      axis.line.y = element_line(color = "black"),
      panel.grid = element_blank(),
      axis.ticks.x = element_line(),
      axis.ticks.y = element_line(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      plot.margin = margin(10, 10, 10, 10)
    )
  
# Plot 1: Relative Abundance (relabund.rpkm)
  p_relabund <- ggplot(df, aes(x = .data[[consolidation.level]], y = relabund.rpkm.mean.normalized, fill = name)) +
    geom_bar(stat = "identity", color = "Dark Blue") +
    scale_y_continuous(limits = c(0, 1.01), expand = c(0, 0)) +
    scale_x_discrete(expand = c(0.09, 0.09)) +
    labs(
      title = relabund_title,
      x = tools::toTitleCase(consolidation.level),
      y = "Relative Abundance of RPKM",
      fill = fill_label
    ) +
    guides(fill = guide_legend(ncol = ncol)) +
    base_theme

  # Plot 2: Raw RPKM abundance
  p_rpkm <- ggplot(df, aes(x = .data[[consolidation.level]], y = rpkm.mean, fill = name)) +
    geom_bar(stat = "identity", color = "Dark Blue") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    scale_x_discrete(expand = c(0.09, 0.09)) +
    labs(
      title = abundance_title,
      x = tools::toTitleCase(consolidation.level),
      y = "Average RPKM",
      fill = fill_label
    ) +
    guides(fill = guide_legend(ncol = ncol)) +
    base_theme
  
  # Combining relabund and abundance plots
  # Remove legend from left plot
  p_relabund_no_legend <- p_relabund + theme(legend.position = "none") + labs(title = relabund_label)
  p_rpkm_with_legend <- p_rpkm + labs(title = abundance_label)

  p_combined <- (p_relabund_no_legend | p_rpkm_with_legend) +
  plot_layout(guides = "collect") +
    plot_annotation(
    title = plot_title,
    theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 10)))
  ) &
  theme(legend.position = "right")
  
  # Save plots in separate folders
  ggsave(file = file.path("consolidated_graphs/1.relabund", paste0("plot_relabund_", consolidation.level, "_", prefix, ".pdf")), 
         plot = p_relabund, width = width, height = height)
  
  ggsave(file = file.path("consolidated_graphs/2.rpkm", paste0("plot_rpkm_", consolidation.level, "_", prefix, ".pdf")), 
         plot = p_rpkm, width = width, height = height)
  
  ggsave(file = file.path("consolidated_graphs/3.combined", paste0("plot_combined_", consolidation.level, "_", prefix, ".pdf")), 
         plot = p_combined, width = width * 1.5, height = height)

  # Return both plots as a named list
  return(list(
    relabund_plot = p_relabund,
    rpkm_plot = p_rpkm,
    combined_plot = p_combined
  ))
}


```

# Make consolidated plots
```{r}
# Create directories if they don't already exist
dir.create("consolidated_graphs/1.relabund", showWarnings = FALSE, recursive = TRUE)
dir.create("consolidated_graphs/2.rpkm", showWarnings = FALSE, recursive = TRUE)
dir.create("consolidated_graphs/3.combined", showWarnings = FALSE, recursive = TRUE)

# by location (for bar plots w/ bar for each loc)
plots_amr.markers.respiratory_influent_by_location <- plot_consolidated_abundance(df_amr.markers.respiratory_influent_by_location, 
                                                                                  "location", w=15, h=10, n=6)
plots_amr.markers.respiratory_solids_by_location <- plot_consolidated_abundance(df_amr.markers.respiratory_solids_by_location, 
                                                                                "location", w=18, h=10, n=7)

plots_amr.markers.urinary_influent_by_location <- plot_consolidated_abundance(df_amr.markers.urinary_influent_by_location, 
                                                                              "location", w=15, h=10, n=6)
plots_amr.markers.urinary_solids_by_location <- plot_consolidated_abundance(df_amr.markers.urinary_solids_by_location, 
                                                                            "location", w=18, h=10, n=7)

plots_micro.respiratory_influent_by_location <- plot_consolidated_abundance(df_micro.respiratory_influent_by_location, 
                                                                            "location", w=10, h=10, n=1)
plots_micro.respiratory_solids_by_location <- plot_consolidated_abundance(df_micro.respiratory_solids_by_location, 
                                                                          "location", w=10, h=10, n=1)

plots_micro.urinary_influent_by_location <- plot_consolidated_abundance(df_micro.urinary_influent_by_location, 
                                                                        "location", w=10, h=10, n=1)
plots_micro.urinary_solids_by_location <- plot_consolidated_abundance(df_micro.urinary_solids_by_location, 
                                                                      "location", w=10, h=10, n=1)


# by type (for 2-bar plots)
plots_amr.markers.respiratory_samples_by_type <- plot_consolidated_abundance(df_amr.markers.respiratory_samples_by_type, 
                                                                             "type", w=16, h=10, n=8)

plots_amr.markers.urinary_samples_by_type <- plot_consolidated_abundance(df_amr.markers.urinary_samples_by_type, 
                                                                         "type", w=16, h=10, n=8)

plots_micro.respiratory_samples_by_type <- plot_consolidated_abundance(df_micro.respiratory_samples_by_type, 
                                                                       "type", w=10, h=10, n=2)

plots_micro.urinary_samples_by_type <- plot_consolidated_abundance(df_micro.urinary_samples_by_type, 
                                                                   "type", w=9, h=10, n=2)


# # View them
# plots_amr.markers.respiratory_samples_by_type$relabund_plot
# plots_amr.markers.respiratory_samples_by_type$rpkm_plot
# plots_amr.markers.respiratory_samples_by_type$combined_plot

```

```{r}

# 1) Respiratory plot (top)
p_respiratory <- ggplot(df_respiratory, aes(x = sample, y = RelAbun, fill = Name)) +
  geom_bar(stat = "identity", color = "Dark Blue", width = 0.9) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1.01), expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0)) +
  labs(
    x = "",           # remove x-axis label "sample"
    y = "Respiratory AMR Relative Abundance",
    fill = "AMR Marker Identity"
  ) +
  theme(
    panel.grid = element_blank(),
    axis.ticks.x = element_blank(),       # remove x-axis ticks
    axis.ticks.y = element_blank(),
    axis.text.x = element_blank(),          # remove sample names
    panel.spacing = unit(0.3, "cm"),
    legend.position = "right",
    plot.margin = margin(t = 0, r = 40, b = 0, l = 5)
  ) +
  facet_wrap(~ Type, scales = "free_x") +
  guides(fill = guide_legend(ncol = 2,
                title.position = "top",    # puts "AMR Marker Identity" above the keys
                label.position = "right")) +
  theme(
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.5, "lines"),
    legend.title = element_text(size = 9)
  )

# 2) Urinary plot (bottom)
p_urinary <- ggplot(df_urinary, aes(x = sample, y = RelAbun, fill = Name)) +
  geom_bar(stat = "identity", color = "Dark Blue", width = 0.9) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1.01), expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0)) +
  labs(
    x = "sample",
    y = "Urinary AMR Relative Abundance",
    fill = ""         # remove legend title for this plot
  ) +
  theme(
    panel.grid = element_blank(),
    axis.ticks.x = element_line(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(0.3, "cm"),
    legend.position = c(1.515, 0.5),
    plot.margin = margin(t = 3, r = 40, b = 0, l = 5)
  ) +
  facet_wrap(~ Type, scales = "free_x") +
  guides(fill = guide_legend(ncol = 3)) +
  theme(
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.5, "lines"),
    legend.title = element_text(size = 9),
    strip.text = element_blank()
  )

p_respiratory <- p_respiratory +
  labs(x = NULL) +  # remove any default x label
  theme(
    axis.title.x = element_blank(),  # remove x title margin
    axis.text.x  = element_blank(),  # remove text
  )

# 3) Combine plots vertically, keeping each legend separate
combined_plot <- p_respiratory / p_urinary + plot_layout(guides = "keep")

pdf("combined_AMR.pdf", width = 9, height = 7)
print(combined_plot)
dev.off()
```

